# DS202.3 – Data Programming with R
# Group C
# Dataset: Clinical tumor sequencing data

### Load required libraries
```{r}
library(readr)
library(dplyr)
library(janitor)
library(stringr)
library(forcats)
library(tidyverse)
library(ggplot2)
library(skimr)
```

### Load the dataset
```{r}
raw_data <- read_tsv("clinical_dataset.tsv")
# View(raw_data)
```

### Make column names consistent
```{r}
cleaned_data <- clean_names(raw_data)
```

### Inspect Structure
```{r}
glimpse(cleaned_data)
# summary(cleaned_data)
```

### Standardize all missing values to true NAs
```{r}
# Standardize various "Unknown" strings to actual NA values
# We apply this only to character columns to avoid type errors with numeric data
cleaned_data <- cleaned_data %>% 
  mutate(across(where(is.character), ~na_if(., "Unknown"))) %>%
  mutate(across(where(is.character), ~na_if(., "NA"))) %>%
  mutate(across(where(is.character), ~na_if(., "[Not Available]")))
```

### Identify and dropp columns with more than 50% missing data
```{r}
threshold <- 0.5
 #Calculate the proportion of NAs for each column
columns_to_keep <- colMeans(is.na(cleaned_data)) < threshold

# Subset the dataframe to keep only valid columns
cleaned_data <- cleaned_data[, columns_to_keep]

print(paste("Columns remaining:", ncol(cleaned_data)))
```

### Address NLP-Derived Fields
```{r}
# Clean specific NLP columns (removing "Yes"/"No" strings and handling conflicts)
# This step creates a 'nlp_site_count' which acts as a proxy for metastatic load
#cleaned_data <- cleaned_data %>%
  # 1. Standardize NLP Yes/No fields to Factors
 # mutate(across(ends_with("nlp"), ~case_when(
  #  . == "Yes" ~ "Yes",
   # . == "No" ~ "No",
    #TRUE ~ NA_character_
#  ))) %>%
 # mutate(across(ends_with("nlp"), as.factor)) %>%

  # 2. Create a feature counting how many sites were flagged by NLP
  # Note: we use rowSums on conditions. We wrap it in a function to handle NAs safely.
  #mutate(nlp_site_count = rowSums(across(starts_with("tumor_site") & ends_with("nlp"), 
                                  #       ~ . == "Yes"), na.rm = TRUE))
```

### Clean and validate primary features
```{r}
cleaned_features <- cleaned_data %>% 
  # --- DEMOGRAPHICS (3 Columns) ---
  mutate(current_age = as.numeric(current_age)) %>% 
  filter(current_age > 0 & current_age < 110) %>% 
  mutate(sex = as.factor(sex)) %>% 
  mutate(race = as.factor(race)) %>% 

  # --- CLINICAL PATHOLOGY (4 Columns) ---
  mutate(cancer_type = as.factor(cancer_type)) %>% 
  mutate(cancer_type_detailed = as.factor(cancer_type_detailed)) %>% 
  mutate(primary_tumor_site = as.factor(primary_tumor_site)) %>% 
  
  # Simplify Stage
  mutate(stage_simplified = case_when(
    grepl("Stage 4|Metastatic|Distant", stage_highest_recorded) ~ "Advanced",
    grepl("Stage 1-3|Localized|Regional", stage_highest_recorded) ~ "Early/Regional",
    TRUE ~ NA_character_ 
  )) %>% 
  mutate(stage_simplified = as.factor(stage_simplified)) %>%

  # --- GENOMIC & MOLECULAR (5 Columns) ---
  # These are crucial for predicting response in sequencing data
  mutate(fraction_genome_altered = as.numeric(fraction_genome_altered)) %>% 
  mutate(mutation_count = as.numeric(mutation_count)) %>% 
  mutate(tmb_nonsynonymous = as.numeric(tmb_nonsynonymous)) %>% 
  mutate(tumor_purity = as.numeric(tumor_purity)) %>% 
  mutate(msi_type = as.factor(msi_type)) %>% # Microsatellite instability impacts treatment

  # --- NLP / HISTORY (2 Columns) ---
  mutate(smoking_history_nlp = as.factor(smoking_history_nlp)) %>% 
  
  # --- TARGET VARIABLE ---
  mutate(response = if_else(overall_survival_status == "0:LIVING", "Responded", "Non-Responded")) %>% 
  mutate(response = as.factor(response))
```

### Final selection of clean features for your analysis
```{r}
final_analysis_df <- cleaned_features %>%
  select(
    # Demographics
    current_age, sex, race,
    # Clinical
    cancer_type, cancer_type_detailed, primary_tumor_site, stage_simplified,
    # Genomics (The "Data Programming" part of the module often looks for these numeric fields)
    fraction_genome_altered, mutation_count, tmb_nonsynonymous, tumor_purity, msi_type,
    # History
    smoking_history_nlp,
    # Target
    response
  ) %>%
  drop_na()

# Check the final dimensions (Should be roughly 14 columns + rows)
dim(final_analysis_df)
summary(final_analysis_df)
```
```{r}
glimpse(final_analysis_df)
```


```{r}
View(final_analysis_df)
```


#EDA 

#Bivariate Analysis
```{r}
ggplot(final_analysis_df, aes(x = current_age)) +
  geom_histogram(bins = 30) +
  labs(
    title = "Distribution of Patient Age",
    x = "Age (Years)",
    y = "Count"
    )

```

```{r}
ggplot(final_analysis_df, aes(x = sex,fill=sex)) +
  geom_bar() +
  labs(
    title = "Gender Distribution",
    x = "Gender",
    y = "Count",
    fill="sex"
  )+
  scale_fill_brewer(palette="Set2")+
  theme_minimal()
```

```{r}
ggplot(data=final_analysis_df,
       aes(x=cancer_type,y=current_age,fill=cancer_type))+
geom_boxplot()+
  labs(title="cancertpe vs current age",x="Cancer_type",y="current_age")
```

```{r}
ggplot(final_analysis_df, aes(x = fct_infreq(cancer_type),fill=sex)) +
  geom_bar(position="fill") +
  coord_flip() +
  labs(
    title = "Distribution of Cancer Types with gender",
    x = "sex",
    y = "Count",
    fill="sex"
  )+
  scale_fill_brewer(palette = "Set2")+ #add some nice colors
  theme_minimal()


```

```{r}
ggplot(final_analysis_df, aes(x = response, fill=response)) +
  geom_bar() +
  labs(
    title = "Responded count",
    x = "Response",
    y = "Count",
    fill="response"
  )+
  scale_fill_brewer(palette="Set2")+
  theme_minimal()
```

```{r}
ggplot(final_analysis_df, aes(x =response, fill=stage_simplified)) +
  geom_bar(position = "fill") +
  labs(title = "Treatment Response by simplified_stage", y = "stage_simplified")
```

```{r}

ggplot(final_analysis_df,
       aes(x = msi_type,
           fill = cancer_type)) +
  geom_bar(position = "fill") +
  labs(
    title = "Treatment Response by Tumor Stage",
    x = "msi_type",
    y = "cancer_type"
  )
```

#Univariate Analysis
```{r}
ggplot(final_analysis_df,
       aes(x=response,
           y = current_age)) +
  geom_boxplot() +
  labs(
    title = "Age Distribution by Treatment Response",
    x = "treatment_outcome",
    y = "Age"
  )
```

```{r}
ggplot(final_analysis_df,
       aes(x = stage_simplified ,
           y = current_age,
           fill = stage_simplified )) +
  geom_boxplot() +
  labs(
    title = "Age vs Tumor Stage by Treatment Response",
    x = "stage_simplified",
    y = "Age (Years)"
  )
```

#check missing values
#class balance check(importent for modeling)
```{r}
colSums(is.na(final_analysis_df))
prop.table(table(final_analysis_df$response))

```

